---
title: "Test_markdown"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




## Testing Rstudio and gg2plot
```{r loading data}
library(ggplot2)
library(plyr)
library(reshape2)

setwd("SMPE_bailly")
my_data = read.csv("Data/Mroz.csv");
str(my_data)

qplot(data=my_data, x=agew, y=wagew, main="Basic qplot")

```
Just a basic qplot here, from which it is hard to see anything. Two phenomenons are still interesting: the rarity of very high wages (not really a surprise), and some wages equal to 0. 
Let's see if we can do better.

```{r explore}
qplot(data=my_data, x=agew, y=wagew, color=work, main="Basic qplot")
```
With colours, it is a little bit better. What is unexpected is the very low number of women who were working. Worst, most of those who participated in the labor-force were not payed! That is strange, for sure, but it is not a mistake (we did not inverted the two), this is really what is in the dataset.

Nevermind, we will see that later. 

```{r wage_and_experience1}

ggplot(data=my_data, aes(x=experience, y=wagew, group=experience)) +geom_boxplot()

```
Well, since x is continuous it is not really nice ...

```{r wage_and_experience2}
exp_cat <- cut(my_data$experience, seq(0, 40, 10))
ggplot(data=my_data, aes(x=exp_cat, y=wagew, group=experience)) +geom_boxplot()
```

```{r wage_and_experience3}
p <- ggplot(data=my_data, aes(x=experience, y=wagew, group=experience)) +geom_boxplot()

p <- p + scale_x_continuous(name = "Years of experience",
                              breaks = seq(0, 40, 5),
                              limits=c(0, 40))
p
```
On the road again, again... But this time, we will eliminate those who don't work

LITTLE TRICK: since it seems work and no work are inverted...
```{r wage_and_experience4}
p <- ggplot(data=my_data[my_data$work=="no", ], aes(x=experience, y=wagew, group=experience)) +geom_boxplot()
p
```
Well, we have something ok here! Not sure we can conclude anything, but still, enough for today.

